Notes

### 2022-08-30 chapters/abstract.tex
Finish up to Ch.6
Revision

\noindent We propose a novel Bayesian approach for functional linear and logistic regression models based on the theory of Reproducing Kernel Hilbert Spaces (RKHS's). These new models build upon the RKHS associated with the covariance function of the underlying stochastic process, and can be viewed as a finite-dimensional approximation to the classical functional regression paradigm. The corresponding functional model (or the functional logistic equation in the case of binary response) is determined by a function living on a dense subspace of the RKHS of interest. By imposing a suitable prior distribution on this space, we can perform data-driven inference via standard Bayes methodology, and the posterior distribution can be estimated through Markov chain Monte Carlo methods. Several prediction strategies derived from this posterior distribution turn out to be competitive against other usual alternatives in both simulated examples and real data sets, including a Bayesian-motivated variable selection procedure.\\

@@

\incomment{add blank page}

>>>>

\noindent This work is framed within the field of Functional Data Analysis, a branch of statistics in which the objects of interest are random functions in a functional space instead of, say, random points in \(\R^p\). Due to the infinite-dimensional nature of the data, the most common \(L^2\)-models for functional linear and logistic regression present some complications that require further simplifications, usually in the form of regularization or dimensionality reduction.

We propose a novel Bayesian approach for functional linear and logistic regression models based on the theory of reproducing kernel Hilbert spaces (RKHS's). These recently-proposed models build upon the RKHS associated with the covariance function of the underlying stochastic process, and can be viewed as a finite-dimensional approximation to the classical functional regression paradigm. The corresponding functional model (or the functional logistic equation in the case of binary response) is determined by a function living on a dense subspace of the RKHS of interest, which has a tractable parametric form. By imposing a suitable prior distribution on this space, we can perform data-driven inference via standard Bayes methodology. The posterior distribution can be estimated through Markov chain Monte Carlo methods, which do not require a complete specification of the posterior density.

We derive several prediction strategies from the approximate posterior distribution, including a Bayesian-motivated variable selection procedure. We show through a comprehensive set of experiments that these methods are competitive against other usual alternatives in terms of predictive performance, both in simulated examples and real data sets. Overall, our proposed model is simple with regard to interpretation and implementation, and has an added flexibility thanks to the ambient Bayesian framework.\\

@@

### 2022-08-30 chapters/appendix-code.tex
Revision

\begin{outcomment}
  Comentar cómo usar el código, las versiones de los paquetes, y los notebooks.

  Contar que son clases a la sklearn y compatibles con ska-fda. Enlazar al github.

  El proceso es reproducible (semilla 2022)

  Se usan 4 cores.

\end{outcomment}

>>>>

\begin{outcomment}
  Aquí irá un pequeño resumen de la estructura del código y los archivos, y un mini-tutorial de cómo usarlo para reproducir los experimentos. No debería ocupar más de una página.

  \begin{itemize}
    \item  Comentar cómo usar el código, las versiones de los paquetes, y los notebooks.
    \item Contar que son clases a la sklearn y compatibles con ska-fda. Enlazar al github.
    \item Se usan 4 cores.
    \item El proceso es reproducible (semilla 2022).
  \end{itemize}
\end{outcomment}

### 2022-08-30 chapters/background-related-work.tex
Revision

 The most common scalar-on-function linear regression model is the classical \(L^2\)-model, widely popularized since the first edition (1997) of the monograph by~\citet{ramsay2005functional}. It can be seen as a generalization of the usual finite-dimensional model, replacing the scalar product in \(\R^p\) for that of the functional space \(L^2[0,1]\):
\begin{equation}\label{eq:l2-linear-model}
Y = \alpha_0 + \dotprod{X}{\beta} + \epsilon = \alpha_0 + \int_0^1 X(t)\beta(t)\, dt + \epsilon,
\end{equation}
where \(\alpha_0\in \R\), \(\epsilon\) is a random error term independent from \(X\) with \(\E [\epsilon]=0\) and \(\E[X(t)\epsilon]=0\) almost everwhere; and the functional slope parameter \(\beta=\beta(\cdot)\) is assumed to be a member of the infinite-dimensional space \(L^2[0, 1]\). A careful rearrangement of~\eqref{eq:l2-linear-model} shows that the model is equivalently expressed as \(\Delta=\K\beta\), where \(\Delta(\cdot)\) is the cross-covariance function of \(X\) and \(Y\), and \(\K\) is the covariance operator of the process \(X\) (to be defined in Section~\ref{sec:rkhs}). This expression is a continuous analog of the normal equations that arise in finite-dimensional regression settings, and since the operator \(\K\) is non-invertible in general, questions of existence and unicity of \(\beta\) need further study (see \citet{cardot2011functional} for more insights).

@@

\begin{theorem}[Karhunen-Loève expansion]
  With the assumptions and notations of Theorem~\eqref{th:mercer}, the centered process \(X=X(t)\) with covariance function \(K\) admits the quadratic-mean representation
  \[
  X(t) = \sum_{j=1}^\infty \zeta_j \phi_j(t),
  \]
  where the convergence is in \(L^2(\Omega)\) and uniform in \(t\). The \(\zeta_j\) are independent zero-mean random variables with \(\E[\zeta_i\zeta_j]=\delta_{ij}\lambda_j\), given explictly by the formula
  \[
    \zeta_j = \int_0^1 X(t)\phi_j(t)\, dt, \quad j\geq 1.
  \]
\end{theorem}

@@

Generally speaking, a MCMC algorithm can be described at time \(t+1\) as \(\Lambda(t+1)=R(\Lambda(t), \xi(t), \pi)\), where \(\Lambda(t)\) is the state of the chain at instant \(t\), \(\pi(x)\) is the objective distribution, and \(\xi(t)\) is a sequence of i.i.d. random variables that represent the random behavior of the chain. With this notation, the property of affine-invariant can be characterized as \citep{goodman2010ensemble}:
 \[
 R(A\lambda+b, \xi(t), \pi_{A,b}) = AR(\lambda, \xi(t), \pi) + b,
 \]
 for all \(A,b\) and \(\lambda\), and for allmost all \(\xi(t)\). This means that if we fix a random generator and run the algorithm twice, one time using \(\pi\) and starting in \(\Lambda(0)\) and a second time using \(\pi_{A,b}\) with initial point \(\Gamma(0)=A\Lambda(0)+b\), then \(\Gamma(t)=A\Gamma(t)+b\) for all \(t\). In \citep{goodman2010ensemble} the authors consider an ensemble of samplers with the affine invariance property. Specifically, they work with a set \(\Lambda=(\Lambda_1, \dots, \Lambda_L)\) of \textit{walkers}, where \(\Lambda_l(t)\) represents an individual chain at time \(t\). At each iteration, an affine-invariant transformation is used to find the next point, which is constructed using the current values of the rest of the walkers (similar to Gibb's algorithm), namely the \textit{complementary ensemble}
 \[
  \Lambda_{-l}(t) = \{\Lambda_1(t+1), \dots, \Lambda_{l-1}(t+1), \Lambda_{l+1}(t), \dots, \Lambda_L(t)\}, \quad l=1,\dots, L.
  \]

@@

From a computational perspective, the Python library \textit{emcee} \footnote{\url{https://emcee.readthedocs.io/en/stable/}} \citep{foreman2013emcee} provides a parallel implementation of this algorithm. The idea is to divide the ensemble \(\Lambda\) into two equally-sized subsets \(\Lambda^{(0)}\) and \(\Lambda^{(1)}\), and then proceed on each iteration in the following alternate fashion:

>>>>

 The most common scalar-on-function linear regression model is the classical \(L^2\)-model, widely popularized since the first edition (1997) of the monograph by~\citet{ramsay2005functional}. It can be seen as a generalization of the usual finite-dimensional model, replacing the scalar product in \(\R^p\) for that of the functional space \(L^2[0,1]\):
\begin{equation}\label{eq:l2-linear-model}
Y = \alpha_0 + \dotprod{X}{\beta} + \epsilon = \alpha_0 + \int_0^1 X(t)\beta(t)\, dt + \epsilon,
\end{equation}
where \(\alpha_0\in \R\), \(\epsilon\) is a random error term independent from \(X\) with \(\E [\epsilon]=0\) and \(\E[X(t)\epsilon]=0\) almost everywhere; and the functional slope parameter \(\beta=\beta(\cdot)\) is assumed to be a member of the infinite-dimensional space \(L^2[0, 1]\). A careful rearrangement of~\eqref{eq:l2-linear-model} shows that the model is equivalently expressed as \(\Delta=\K\beta\), where \(\Delta(\cdot)\) is the cross-covariance function of \(X\) and \(Y\), and \(\K\) is the covariance operator of the process \(X\) (to be defined in Section~\ref{sec:rkhs}). This expression is a continuous analog of the normal equations that arise in finite-dimensional regression settings, and since the operator \(\K\) is non-invertible in general, questions of existence and unicity of \(\beta\) need further study (see \citet{cardot2011functional} for more insights).

@@

\begin{theorem}[Karhunen-Loève expansion]
  With the assumptions and notations of Theorem~\eqref{th:mercer}, the centered process \(X=X(t)\) with covariance function \(K\) admits the quadratic-mean representation
  \[
  X(t) = \sum_{j=1}^\infty \zeta_j \phi_j(t),
  \]
  where the convergence is in \(L^2(\Omega)\) and uniform in \(t\). The \(\zeta_j\) are independent zero-mean random variables with \(\E[\zeta_i\zeta_j]=\delta_{ij}\lambda_j\), given explicitly by the formula
  \[
    \zeta_j = \int_0^1 X(t)\phi_j(t)\, dt, \quad j\geq 1.
  \]
\end{theorem}

@@

Generally speaking, a MCMC algorithm can be described at time \(t+1\) as \(\Lambda(t+1)=R(\Lambda(t), \xi(t), \pi)\), where \(\Lambda(t)\) is the state of the chain at instant \(t\), \(\pi(x)\) is the objective distribution, and \(\xi(t)\) is a sequence of i.i.d. random variables that represent the random behavior of the chain. With this notation, the property of affine-invariant can be characterized as \citep{goodman2010ensemble}:
 \[
 R(A\lambda+b, \xi(t), \pi_{A,b}) = AR(\lambda, \xi(t), \pi) + b,
 \]
 for all \(A,b\) and \(\lambda\), and for almost all \(\xi(t)\). This means that if we fix a random generator and run the algorithm twice, one time using \(\pi\) and starting in \(\Lambda(0)\) and a second time using \(\pi_{A,b}\) with initial point \(\Gamma(0)=A\Lambda(0)+b\), then \(\Gamma(t)=A\Gamma(t)+b\) for all \(t\). In \citep{goodman2010ensemble} the authors consider an ensemble of samplers with the affine invariance property. Specifically, they work with a set \(\Lambda=(\Lambda_1, \dots, \Lambda_L)\) of \textit{walkers}, where \(\Lambda_l(t)\) represents an individual chain at time \(t\). At each iteration, an affine-invariant transformation is used to find the next point, which is constructed using the current values of the rest of the walkers (similar to Gibb's algorithm), namely the \textit{complementary ensemble}
 \[
  \Lambda_{-l}(t) = \{\Lambda_1(t+1), \dots, \Lambda_{l-1}(t+1), \Lambda_{l+1}(t), \dots, \Lambda_L(t)\}, \quad l=1,\dots, L.
  \]

@@

From a computational perspective, the Python library \textit{emcee}\footnote{\url{https://emcee.readthedocs.io/en/stable/}} \citep{foreman2013emcee} provides a parallel implementation of this algorithm. The idea is to divide the ensemble \(\Lambda\) into two equally-sized subsets \(\Lambda^{(0)}\) and \(\Lambda^{(1)}\), and then proceed on each iteration in the following alternate fashion:

### 2022-08-30 chapters/bayesian-functional-regression.tex
Revision

In this chapter we present the precise functional models and Bayesian methodologies explored in this work. The RKHS-based functional models under consideration \citep[see][]{berrendero2018functional, berrendero2019rkhs} are those obtained by substituting the functional parameter \(\beta\in L^2[0,1]\) for \(\alpha \in \Hcal(K)\), replacing also the scalar product \(\dotprod{X}{\beta}\) for \(\dotprod{X}{\alpha}_K\) in the \(L^2\)-models~\eqref{eq:l2-linear-model} and~\eqref{eq:l2-logistic-model}. However, to further simplify things we will follow a parametric approach and suppose that \(\alpha\) is in fact a member of the dense subspace \(\Hcal_0(K)\) defined in~\eqref{eq:h0}, i.e.:
\begin{equation}\label{eq:alpha_h0}
\alpha(\cdot) = \sum_{j=1}^p \beta_j K(t_j, \cdot), \text{ for some } p \in \N, \ \beta_j \in \R \text{ and } t_j \in [0,1].
\end{equation}
Moreover, as we said before, with a slight abuse of notation we will understand the expression \(\dotprod{x}{\alpha}_K\) as \(\Psi_x(\alpha)\), where \(x=X(\omega)\) and \(\Psi_x\) is Loève's isometry. Hence, taking into account that \(\Psi_X(K(t, \cdot)) = X(t)\) by definition (see~\eqref{eq:loeves-isometry}), we can write \(\dotprod{x}{\alpha}_K \equiv \sum_{j=1}^p \beta_j x(t_j)\) when \(\alpha\) is as in \eqref{eq:alpha_h0}.

In this way we get a simpler, finite-dimensional approximation of the functional RKHS model, which we argue reduces the overall complexity of the model while still capturing most of the relevant information. When it comes to parameter estimation, a direct optimization of some loss function would probably require a tailored algorithm that took into account the continuous nature of the times \(t_j\). Indeed, such an idea is explored in \citet{berrendero2018functional} for the logistic regression case, where the authors propose a ``greedy max-max'' method reminescent of the EM algorithm that alternates between estimating the coefficients and the time instants through a maximum likelihood approach.

@@

\subsection*{The bayesian approach: prior and posterior}

>>>>

In this chapter we present the precise functional models and Bayesian methodologies explored in this work. The RKHS-based functional models under consideration \citep[see][]{berrendero2018functional, berrendero2019rkhs} are those obtained by substituting the functional parameter \(\beta\in L^2[0,1]\) for \(\alpha \in \Hcal(K)\), replacing also the scalar product \(\dotprod{X}{\beta}\) for \(\dotprod{X}{\alpha}_K\) in the \(L^2\)-models~\eqref{eq:l2-linear-model} and~\eqref{eq:l2-logistic-model}. However, to further simplify things we will follow a parametric approach and suppose that \(\alpha\) is in fact a member of the dense subspace \(\Hcal_0(K)\) defined in~\eqref{eq:h0}, i.e.:
\begin{equation}\label{eq:alpha_h0}
\alpha(\cdot) = \sum_{j=1}^p \beta_j K(t_j, \cdot), \text{ for some } p \in \N, \ \beta_j \in \R \text{ and } t_j \in [0,1].
\end{equation}
Moreover, as we said before, with a slight abuse of notation we will understand the expression \(\dotprod{x}{\alpha}_K\) as \(\Psi_x(\alpha)\), where \(x=X(\omega)\) and \(\Psi_x\) is Loève's isometry. Hence, taking into account that \(\Psi_X(K(t, \cdot)) = X(t)\) by definition (see~\eqref{eq:loeves-isometry}), we can write \(\dotprod{x}{\alpha}_K \equiv \sum_{j=1}^p \beta_j x(t_j)\) when \(\alpha\) is as in~\eqref{eq:alpha_h0}.

In this way we get a simpler, finite-dimensional approximation of the functional RKHS model, which we argue reduces the overall complexity of the model while still capturing most of the relevant information. When it comes to parameter estimation, a direct optimization of some loss function would probably require a tailored algorithm that took into account the continuous nature of the times \(t_j\). Indeed, such an idea is explored in \citet{berrendero2018functional} for the logistic regression case, where the authors propose a ``greedy max-max'' method reminiscent of the EM algorithm that alternates between estimating the coefficients and the time instants through a maximum likelihood approach.

@@

### 2022-08-30 chapters/conclusions.tex
Revision

\begin{outcomment}
  - Destacar también las ventajas de lo bayesiano: flexibilidad, incorporación de información con distintas priors, etc. En la introducción, conclusiones, experimentos, ..
- Destacar interpretabilidad en los experimentos: e.g. un valor de beta=3.4 se puede interpretar fácilmente en un caso concreto. En general, nuestros modelos tienen menor dimensión.
- Logistic un poco mejor que linear?
- Tiempo de ejecución alto (característico de los algoritmos MCMC), a cambio mucha sencillez.
- CV ha sido pequeño. Con más parámetros, probablemente mejores resultados. También flexibilidad para cambiar las prioris.
- We do not advise to use the mean as a summary statistic.
\end{outcomment}

@@

\begin{outcomment}
  - Mejorar pymc.
  - consistencia de la posteriori, robustez de la priori
  - Otras prioris

\end{outcomment}

>>>>

\begin{outcomment}
  Algo que no he comentado en ninguna parte es que el tiempo de ejecución de nuestros modelos es bastante alto (del orden de 30-40 segundos por ejecución) frente al de los algoritmos de comparación (unos pocos segundos en general). Como es algo negativo, no sé muy bien cómo ponerlo, si comentarlo de pasada aquí en las conclusiones, o dedicarle un poco más de espacio en el capítulo de experimentos.\\
\end{outcomment}

\begin{outcomment}
  Algunos comentarios generales:
  \begin{itemize}
    \item Destacar las ventajas de lo bayesiano: flexibilidad, incorporación de información con distintas priors, etc.
    \item Destacar la interpretabilidad y sencillez de nuestro modelo. En general, conseguimos resultados competitivos usando modelos con baja dimensión.
    \item Los resultados en regresión logística parecen en media un poco mejores que en regresión lineal.
    \item No se recomienda usar la media como estadístico de resumen de la distribución a posteriori.
    \item La búsqueda de hiperparámetros en CV no ha sido demasiado grande. Probablemente con más recursos computacionales se podrían conseguir mejores resultados.
  \end{itemize}
\end{outcomment}

@@

\begin{outcomment}
  \begin{itemize}
    \item Ahondar en la relación entre RKHS y regresión funcional.
    \item Experimentar con otros algoritmos de MCMC (en particular, reversible-jump MCMC), y con otros paquetes (pymc).
    \item Considerar otras distribuciones a priori.
    \item Derivar algunos resultados de consistencia y/o robustez de la posteriori.
    \item Extender el modelo lineal a un modelo GLM.
  \end{itemize}
\end{outcomment}

### 2022-08-30 chapters/experiments.tex
Revision

Lastly, since the use of MCMC algorithms introduces a source of stochasticity in the prediction procedure, we independently repeat the whole process 10 times (each with a different train/test configuration), and average the results across these executions. The metrics used to evaluate the perfomance of the models are the RMSE for linear regression and the accuracy for logistic regression. Recall that
\[
\operatorname{RMSE}(\{\hat y_i\}) = \sqrt{\frac{1}{n'}\sum_{i=1}^{n'} (y_i - \hat y_i)^2}
\]
and
\[
\operatorname{Accuracy}(\{\hat y_i\}) = \frac{1}{n'} \sum_{i=1}^{n'} \I(y_i = \hat y_i).
\]

@@

We consider a set of functional regressors common to linear and logistic regression problems. They are 4 Gaussian processes (GP's), each with a different kernel function:
\begin{enumerate}
  \item A Brownian motion \(K_1(t,s)=\min\{t,s\}\).
  \item A fractional Browmian motion \(K_2(t,s)=1/2(s^{2H} + t^{2H} - |t-s|^{2H})\), with Hurst parameter \(H=0.8\).
  \item An Ornstein-Uhlenbeck process \(K_3(t,s)=e^{-|t-s|}\).
  \item A Gaussian kernel \(K_4(t,s)=e^{-(t-s)^2/2\nu^2}\), with \(\nu=0.2\).
\end{enumerate}
Also, when applicable, we fix a variance \(\sigma^2=0.5\) for the error terms \(\epsilon\).

@@

\paragraph{Logistic regression data sets.} Again we consider three different types of data sets,  with a common value of \(\alpha_0=-0.5\).
\begin{itemize}
  \item Four logistic RKHS models with the same functional parameter as in the linear regression case (one for each GP). Specifically,
  \[
  \P(Y=1\mid X) = \frac{1}{e^{0.5 +5X(0.1) - X(0.4) - 10X(0.8)}}.
  \]
  \item Four logistic \(L^2\)-models with the same coefficient function as in the linear regression case.
  \item A ``mixture'' model based on non-GP regressors, in which we mix regressors from two different GP's and label them according to their origin. Firstly, we consider a homoscedastic case to distinguish between a standard Browmian motion and a Brownian motion with a mean function that is zero until \(t=0.5\), and then becomes \(m(t)=0.75t\). Secondly, we consider a heteroscedastic case to distinguish between a standard Brownian motion and a Brownian motion with variance \(2\), that is, with kernel \(K(t,s)=2\min\{t,s\}\).
\end{itemize}
In this case there are again three real data sets. The first one is a subset of the Medflies data set \citep{carey1998relationship}, consisting on samples of the number of eggs laid daily by 534 flies over 30 days, trying to predict if their longevity is high or low. The second one is the Berkeley Growth Study data set \citep{tuddenham1954physical}, which records the height of 54 girls and 39 boys over 31 different points in their lives. Finally, we have selected a subset of the Phoneme data set \citep{hastie1995penalized}, based on 200 digitized speech frames over 128 equispaced points to predict the phonemes ``aa'' and ``ao''.

@@

We have included a fairly comprehensive suite of comparison algorithms, chosen among the most common methods used in machine learning and FDA. There are finite-dimensional models that work on the discretized data, variable selection/dimension reduction procedures and purley functional methods.

\paragraph{Linear regression comparison algorithms.} We consider the following methods.
\begin{itemize}
  \item Manual. Dummy variable selection method with a pre-specified number of components (equispaced on \([0, 1]\)).
  \item Lasso. Linear least squares with \(l^1\) regularization.
  \item Ridge. Linear least squares with \(l^2\) regularization.
  \item PLS. Partial least squares for dimension reduction.
  \item PCA. Principal component analysis for dimension reduction.
  \item PLS1. Partial least squares regression \citep[e.g.][]{wegelin2000survey}.
  \item APLS. Functional partial least squares regression proposed by \citet{delaigle2012methodology}.
  \item RMH. Recursive maxima hunting variable selection method proposed by \citet{torrecilla2016feature}.
  \item FLin. Functional \(L^2\) linear regression model with fixed basis expansion.
  \item FPLS1. Functional PLS regression through basis expansion, implemented as in \citet{aguilera2010using}.
  \item FPCA. Functional principal component analysis.
\end{itemize}

\paragraph{Logistic regression comparison algorithms.} All the variable selection and dimension reduction algorithms from above are also considered in this case.
\begin{itemize}
  \item Log. Standard multiple logistic regression with \(l^2\) regularization.
  \item LDA. Linear discriminant analysis.
  \item QDA. Quadratic discriminant analysis.
  \item RKVS. RKHS-based variable selection method proposed in \citet{berrendero2018use}.
  \item APLS. Functional PLS used as a dimension reduction method, as proposed in \citet{delaigle2012achieving} in combination with the nearest centroid algorithm.
  \item FLog. Functional RKHS-based logistic regression algorithm proposed in \citet{berrendero2018functional}.
  \item FLDA. Implementation of the so-called functional linear discriminant analysis in \citet{preda2007pls}.
  \item MDC. Maximum depth classifier \citep[e.g.][]{ghosh2005maximum}.
  \item FKNN. Functional K-nearest neighbors classifier with the \(L^2\)-distance.
  \item FNC. Functional nearest centroid classifier with the \(L^2\)-distance.
\end{itemize}

@@

In Figure~\ref{fig:reg_emcee_l2} we see the results for the case with an underlying \(L^2\)-model, which would be our most direct competitor. In this case the outcome is satisfactory, since for the most part our models are on a par with the rest, even beating other methods that were designed with the \(L^2\)-model in mind. Moreover, whenever one of our models has a higher RMSE, the difference is pretty small in general.
\begin{figure}[ht!]
  \centering
  \includegraphics[width=\textwidth]{reg_emcee_l2}
  \caption{RMSE of predictors for simulated GP data that obeys an underlying \(L^2\) model (lower is better).}\label{fig:reg_emcee_l2}
\end{figure}

@@

Figure~\ref{fig:reg_emcee_real} shows the results for the real data sets. In these data sets there is a substantial difference in perfomance between some of our methods and the reference algorithms. However, the predict-then-summarize approach (represented as \textit{posterior\_mean}) seems to work quite well, always near the mean RMSE of all the comparison algorithms. Moreover, our two-stage methods seem to perform better than our other methods in these cases, scoring again very close to the mean of the reference methods.

@@

As we can see, one of the coefficients has gone to zero to account for the overspecification of the model, while the other two have stabilized very close to the true parameters. The same goes for the time instants, except that in this case there is no default value to represent that a component is unused, so the time corresponding to the null coefficient oscillates back and forth. The estimated function, based for example in the mode of the posterior distributions, will not perfect, essentially because of the noise in the response. But it should be close to the true parameter function \(\alpha(t)=-5K(t, 0.1) + 10K(t, 0.8)\), providing a good predictive performance.

In contrast, if we consider now a model with \(p=4\) with the same data, we might obtain posterior distributions such as the ones in Figure~\ref{fig:beta_trace_4}. In this situation two components should go to zero, but that is no longer the case. For example, while the green component has a high density around 0, it also has a considerable mass around 10, effectively ``competing'' with the red component. This is a manifestation of the label switching issue, caused in this case by an excessive number of degrees of freedom in the model.

@@

Apart from all the experiments described above, we performed several tests with different configurations and alternatives. Although these trials did not produce better results in general, we mention here some of them to show how the model could be easily extended. In fact, many of these modifications, if not all, are still available in the provided code and can be enabled through specific parameters.

The first thing tried were cases in which the parameter \(p\) was part of the Bayesian model, with a categorical prior distribution. As explained in Chapter~\ref{ch:model-choice} this approach led to many complications and generally poor results. However, the idea of letting the model select how many components it needs to have based on the data is an interesting one.

>>>>

Lastly, since the use of MCMC algorithms introduces a source of stochasticity in the prediction procedure, we independently repeat the whole process 10 times (each with a different train/test configuration), and average the results across these executions. The metrics used to evaluate the performance of the models are the RMSE for linear regression and the accuracy for logistic regression. Recall that
\[
\operatorname{RMSE}(\{\hat y_i\}) = \sqrt{\frac{1}{n'}\sum_{i=1}^{n'} (y_i - \hat y_i)^2}
\]
and
\[
\operatorname{Accuracy}(\{\hat y_i\}) = \frac{1}{n'} \sum_{i=1}^{n'} \I(y_i = \hat y_i).
\]

@@

We consider a set of functional regressors common to linear and logistic regression problems. They are 4 Gaussian processes (GP's), each with a different kernel function:
\begin{enumerate}
  \item A Brownian motion \(K_1(t,s)=\min\{t,s\}\).
  \item A fractional Brownian motion \(K_2(t,s)=1/2(s^{2H} + t^{2H} - |t-s|^{2H})\), with Hurst parameter \(H=0.8\).
  \item An Ornstein-Uhlenbeck process \(K_3(t,s)=e^{-|t-s|}\).
  \item A Gaussian kernel \(K_4(t,s)=e^{-(t-s)^2/2\nu^2}\), with \(\nu=0.2\).
\end{enumerate}
Also, when applicable, we fix a variance \(\sigma^2=0.5\) for the error terms \(\epsilon\).

@@

\paragraph{Logistic regression data sets.} Again we consider three different types of data sets,  with a common value of \(\alpha_0=-0.5\).
\begin{itemize}
  \item Four logistic RKHS models with the same functional parameter as in the linear regression case (one for each GP). Specifically,
  \[
  \P(Y=1\mid X) = \frac{1}{e^{0.5 +5X(0.1) - X(0.4) - 10X(0.8)}}.
  \]
  \item Four logistic \(L^2\)-models with the same coefficient function as in the linear regression case.
  \item A ``mixture'' model based on non-GP regressors, in which we mix regressors from two different GP's and label them according to their origin. Firstly, we consider a homoscedastic case to distinguish between a standard Brownian motion and a Brownian motion with a mean function that is zero until \(t=0.5\), and then becomes \(m(t)=0.75t\). Secondly, we consider a heteroscedastic case to distinguish between a standard Brownian motion and a Brownian motion with variance \(2\), that is, with kernel \(K(t,s)=2\min\{t,s\}\).
\end{itemize}
In this case there are again three real data sets. The first one is a subset of the Medflies data set \citep{carey1998relationship}, consisting on samples of the number of eggs laid daily by 534 flies over 30 days, trying to predict if their longevity is high or low. The second one is the Berkeley Growth Study data set \citep{tuddenham1954physical}, which records the height of 54 girls and 39 boys over 31 different points in their lives. Finally, we have selected a subset of the Phoneme data set \citep{hastie1995penalized}, based on 200 digitized speech frames over 128 equispaced points to predict the phonemes ``aa'' and ``ao''.

@@

We have included a fairly comprehensive suite of comparison algorithms, chosen among the most common methods used in machine learning and FDA. There are finite-dimensional models that work on the discretized data, variable selection/dimension reduction procedures and purely functional methods.

\paragraph{Linear regression comparison algorithms.} We consider the following methods.
\begin{itemize}
  \item Manual.\hspace{.3em} Dummy variable selection method with a pre-specified number of components (equispaced on \([0, 1]\)).
  \item Lasso.\hspace{.3em} Linear least squares with \(l^1\) regularization.
  \item Ridge.\hspace{.3em} Linear least squares with \(l^2\) regularization.
  \item PLS.\hspace{.3em} Partial least squares for dimension reduction.
  \item PCA.\hspace{.3em} Principal component analysis for dimension reduction.
  \item PLS1.\hspace{.3em} Partial least squares regression \citep[e.g.][]{wegelin2000survey}.
  \item APLS.\hspace{.3em} Functional partial least squares regression proposed by \citet{delaigle2012methodology}.
  \item RMH.\hspace{.3em} Recursive maxima hunting variable selection method proposed by \citet{torrecilla2016feature}.
  \item FLin.\hspace{.3em} Functional \(L^2\) linear regression model with fixed basis expansion.
  \item FPLS1.\hspace{.3em} Functional PLS regression through basis expansion, implemented as in \citet{aguilera2010using}.
  \item FPCA.\hspace{.3em} Functional principal component analysis.
\end{itemize}

\paragraph{Logistic regression comparison algorithms.} All the variable selection and dimension reduction algorithms from above are also considered in this case.
\begin{itemize}
  \item Log.\hspace{.3em} Standard multiple logistic regression with \(l^2\) regularization.
  \item LDA.\hspace{.3em} Linear discriminant analysis.
  \item QDA.\hspace{.3em} Quadratic discriminant analysis.
  \item RKVS.\hspace{.3em} RKHS-based variable selection method proposed in \citet{berrendero2018use}.
  \item APLS.\hspace{.3em} Functional PLS used as a dimension reduction method, as proposed in \citet{delaigle2012achieving} in combination with the nearest centroid algorithm.
  \item FLog.\hspace{.3em} Functional RKHS-based logistic regression algorithm proposed in \citet{berrendero2018functional}.
  \item FLDA.\hspace{.3em} Implementation of the so-called functional linear discriminant analysis in \citet{preda2007pls}.
  \item MDC.\hspace{.3em} Maximum depth classifier \citep[e.g.][]{ghosh2005maximum}.
  \item FKNN.\hspace{.3em} Functional K-nearest neighbors classifier with the \(L^2\)-distance.
  \item FNC.\hspace{.3em} Functional nearest centroid classifier with the \(L^2\)-distance.
\end{itemize}

@@

In Figure~\ref{fig:reg_emcee_l2} we see the results for the case with an underlying \(L^2\)-model, which would be our most direct competitor. In this case the outcome is satisfactory, since for the most part our models are on a par with the rest, even beating other methods that were designed with the \(L^2\)-model in mind. Moreover, whenever one of our models has a higher RMSE, the difference is pretty small in comparison.
\begin{figure}[ht!]
  \centering
  \includegraphics[width=\textwidth]{reg_emcee_l2}
  \caption{RMSE of predictors for simulated GP data that obeys an underlying \(L^2\) model (lower is better).}\label{fig:reg_emcee_l2}
\end{figure}

@@

Figure~\ref{fig:reg_emcee_real} shows the results for the real data sets. In these data sets there is a substantial difference in performance between some of our methods and the reference algorithms. However, the predict-then-summarize approach (represented as \textit{posterior\_mean}) seems to work quite well, always scoring near the mean RMSE of all the comparison algorithms. Moreover, our two-stage methods seem to perform better than our other methods in these cases, scoring again very close to the mean of the reference methods.

@@

As we can see, one of the coefficients has gone to zero to account for the overspecification of the model, while the other two have stabilized very close to the true parameters. The same goes for the time instants, except that in this case there is no default value to represent that a component is unused, so the time corresponding to the null coefficient oscillates back and forth. The estimated function (based for example in the mode of the posterior distributions) will not be perfect, essentially because of the noise in the response. But it should be close to the true parameter function \(\alpha(t)=-5K(t, 0.1) + 10K(t, 0.8)\), providing a good predictive performance.

In contrast, if we consider now a model with \(p=4\) with the same data, we might obtain posterior distributions such as the ones in Figure~\ref{fig:beta_trace_4}. In this situation two coefficients should go to zero, but that is no longer the case. For example, while the green component has a high density around 0, it also has a considerable mass around 10, effectively ``competing'' with the red component. This is a manifestation of the label switching issue, caused in this case by an excessive number of degrees of freedom in the model.

@@

Apart from all the experiments described above, we performed several tests with different configurations and alternatives. Although these trials did not produce better results in general, we mention some of them here to show how the model could be easily extended. In fact, many of these modifications, if not all, are still available in the provided code and can be enabled through specific parameters.

The first thing we tried was to make the parameter \(p\) part of the Bayesian model, imposing a categorical prior distribution on it. As explained in Chapter~\ref{ch:model-choice} this approach led to many complications and generally poor results. However, the idea of letting the model select how many components it should have based on the data is an interesting one, and in some situations this could greatly improve the results of variable selection procedures based on other model selection criteria.

### 2022-08-30 chapters/introduction.tex
Revision

% \begin{center}
%   \begin{minipage}{0.5\textwidth}
%     \begin{small}
%       \emph{Quote}
%     \end{small}
%   \end{minipage}
%   \vspace{0.5cm}
% \end{center}

@@

\begin{enumerate}[1.]
  \item To perform a brief but thorough literature review that contextualizes functional linear and logistic regression in statistics and machine learning, exploring the predominant models and techniques.
  \item To propose a novel functional model for linear and logistic regression that builds on existing work and focuses on simplicity, both in terms of interpretation and implementation.
  \item To describe a general Bayesian approach for parameter estimation within the suggested model.
  \item To introduce the tools needed to specify and implement the proposed model, mainly reproducing kernel Hilbert spaces and Markov chain Monte Carlo methods.
  \item To carry out a comprehensive experimental study to test these new models and compare them to existing methods, both in simulations and in real-world scenarios.
\end{enumerate}

>>>>

@@

\begin{enumerate}[1.]
  \item To perform a brief but thorough literature review that contextualizes functional linear and logistic regression in statistics and machine learning, exploring the predominant models and techniques.
  \item To propose a novel RKHS-based functional model for linear and logistic regression that builds on existing work and focuses on simplicity, both in terms of interpretation and implementation.
  \item To describe and implement a general Bayesian approach for parameter estimation within the suggested model.
  \item To introduce the tools needed to specify and implement the proposed model, mainly reproducing kernel Hilbert spaces and Markov chain Monte Carlo methods.
  \item To carry out a comprehensive experimental study to test these new models and compare them to existing methods, both in simulations and in real-world scenarios.
\end{enumerate}
